Species motif participation provides unique information about species risk of extinction

### 2022-05-24 manuscript/response_PRSB_round2.tex

\them
Second, as reviewer 2 argues there is scope for including a figure about the motifs in the context of their food web to facilitate the understanding about their structure and ecological meaning. The top of Figure 1 helps, but there is room to expand.

\us
Anna E

\them
Last, the authors generate what they argue are "a range of plausible network architectures".  But there is never any comparison of these architecture to empirical networks.  Nor is there any calculation of the key motif parameters on empirical data (for example the x-axis of figure 1). It will expand the audience for this manuscript greatly to include some empirical data to help empiricists understand for example in well-studied systems, what is the proportion of these different motifs.  This comparison does not need to be exhaustive, but it will serve two main roles in the manuscript: (1) provide support for the assertion that the authors have generated "a range of plausible network architectures" and (2) allow empiricists to understand how these simulations relate to real ecosystems.

\us
Alyssa - this week (or bother me why not)
Webs referenced in the niche model vs empirical motifs paper

@@

Comments to the Author(s)
Thank you for the opportunity to review this paper. I found it very interesting and appreciated the comparison between motifs and more “classic” network metrics. I have a couple major concerns which I do not anticipate being that difficult to address, and some smaller comments.

\us
We thank the reviewer for the thorough engagement with our work, the encouraging words, and the constructive criticisms.

\them
First, I think the introduction could be clearer. I think it spent too much time discussing bottom up effects, which I appreciate because that is one of the major limitations to this approach, however, I think some of the text could be better spent setting up some of the additional questions of the paper and clarifying various terminology (e.g. local, meso, and global scale gets thrown about, but is a bit of a surprise).

\us
The introduction has been re-written, and is hopefully clearer now regarding terminology and the questions asked. Additionally, we have now included a glossary in \emph{Appendix S11}

@@

\them
Second, I found the analysis, particularly the divvying up into 5 parts rather circular. We know that connectance and species richness has a large impact on both motif expression and on persistence. Instead of multiple separate analyses, I think these can all be included in one large model.

\us - Alyssa: S and C important but not main focus, so want to address them and then move on to more detailed things.
a written response here (and something brief in the paper)

@@

\them
Moreover, I would be interested in an analysis which pits trophic level and in-degree against the motifs. In particular, collecting data on trophic level is much easier than collecting entire food web data for motif roles.

\us
Additional analysis - maybe use motifs to explain differences between species with the same degree, trophic level? (try explanation first)
if you have TL and degree you also have the motif profile of a species (disagree that trophic level is "easier" - once the network is characterised all three role types are available) put in the intro as well

@@

\us See above. The only methodological tie between the two manuscripts was when simulating the networks. The underlying software used is peer-reviewed~\citep{Delmas2017} and has been used in several other published studies, as well as the pre-print originally referenced (currently under review). To avoid the mis-perception that we rely on unproven methods, we now present the full network-generation methods in an Appendix.

\them
Smaller comments

Line 65 – 70: I agree that in-degree and trophic level may be coarse description of a species’ role within its food-web. However, the examples provided do not convince me that species’ roles based on motifs are a more nuanced way of describing a species’ role within its food web since the two examples provided refer to the influence disturbance may have on a basal resource – an influence that I’m not convinced a species’ role based on motifs will impact.  
I do think that species’ roles based on motifs are a more nuanced way of describing species’ roles, and I am interested in the central question of this paper – however, this rationale would not convince me if I was unfamiliar with motifs/species roles/finer scale network descriptors.  As this is a central tenet of the paper I think the argument needs to be constructed better. Furthermore, just because a metric is coarse does not mean it is not an effective predictor – and when applying theoretical/computational results to empirical systems sometimes coarse is better. It is easier to measure the trophic position of a species using stable isotopes and approximate extinction probabilities than it is to construct an entire food web.

\us Degree is better than TL and then you still need the whole web - mention limits of stable isotopes (stable isotope guy maybe?) ... Address this in the new (conceptual?) figure. Show motif profiles that should experience strong and weak effects of bottom-up disturbance

@@

\them
Line 82-84: This sentence doesn’t quite make sense to me.

\us
This sentence was removed when re-writing the introduction.

@@

\them
Line 86: is this “focal” species the consumer mentioned in line 85?

\us
Rephrased, yes it is the same species, i.e. do the extinction risk of a consumer depend upon the motifs in which it participates in?

@@

\them
Line 86: the impact of the motif profile of the full food web on the consumer’s extinction risk has come as a bit of a surprise to me. I think this could have been introduced better in the introduction – not extensively because I think it is probably a secondary question, but at this point I have no hypotheses for why I would even expect the motif profile of the full food web to out perform the motifs in which a focal species participates.

\us
We have now revised the introduction to introduce the motif profile of the whole network and its potential relevance to species-level extinction risk [[check]] and hope that the purpose of these analyses is now clearer.
[[edit to be nicer]]We did not expect that the motif profile of a full food web to outperform the impact of a species motif participation, instead, this was done to investigate whether the full profile of a web would give some additional information, in "average" terms.

@@

\them
Line 87 – 89: similar to comment above – I think the idea of global and local food web properties has not been introduced at all, nor the classic food web properties (which ones are being measured? Only the four specific motifs thought to contribute to stability are mentioned).

\us We have revised the introduction to more clearly introduce these concepts.

@@

\them
Line 92 – 95: this seems to be a flaw to the Bayesian network simulations – we know that mutual predation happens in food webs – especially if the food web isn’t explicitly size structured, and when nested within the broader food web context these feedback loops may have stabilizing, or destabilizing effects.

\us This is now re-written, and we describe this limitation of the Bayesian network framework in the Methods section [[might need to make this more explicit in the methods section, it is now mentioned a bit here and there, also add references to Allesina lab papers if we have some additional?]]

@@

\us To the extent that species in higher-connectance networks tend to participate in more motifs, this normalization does address connectance, however that is not its main purpose and it has been tested whether normalizing motif roles in this way fully controls for connectance.
Since this normalization does not remove correlations between degree and motif participation (as discussed in~\citealp[]{Cirtwill2021_inprep}), we are not confident in making such a claim.
Instead, to control for connectance explicitly, we include it (and species richness) in a random effect in all regressions where it is not a main effect. 
We take this approach because re-evaluating the known effects of connectance and species richness is not a major theme in the manuscript.
We are more interested in comparing extinction risk for species within the same food web (where connectance is constant) than between food webs (where connectance, species richness, and --in empirical systems-- local conditions will have large effects).
We have lightly revised the methods section to make this focus clearer and to state that, while our motif role normalization does not control for degree or connectance, we do explicitly consider how degree and connectance (as well as species richness and trophic level) may be related.

    ``Note, however, that this normalization does not necessarily control for differences due to degree or connectance as high- and low-degree species (or species in more- and less-connected networks) may also tend to participate in different types of motifs.
    We address the potential associations between motif participation, degree, and connectance explicitly below.''

\them
Line 182: what is network mean persistence? Species persistence is the probability of species persistence referred to in the methods? Be explicit. Also, what are the “other network properties”?

\us
Added clarifications to to point 3 and 5 where we had "network mean persistence" and "other network properties", please check if ok.

@@

\them
Line 192: the idea of disturbance level came as a surprise – a careful reading of the methods pinpointed where you discuss this, however it is not discussed in the statistical analysis part, or in the main objectives of the paper at the end of the introduction.

\us This concept is now introduced in the introduction [[but should we discuss it a bit more also in the statistical analysis? we mention the different disturbance levels in the end of the methods section Calculating persistence]]

@@

\them
Line 197 – 199: This sounds more statistically rigorous here than it does in the appendix where these are just visually compared.

\us We have added the word ``visually'' to the main text.

\them
Line 201: I have no idea what local and meso-scale refer to. Likewise global.

\us These terms are explained at their first use in the introduction or methods. To avoid confusion, we also add a glossary in \emph{Appendix S11} so that readers can easily look up these terms. We have also revised the paragraph starting on line 201 to include parenthetical definitions and clarify the scale which we are considering.

### 2022-05-24 manuscript/reorder_main_text.tex
Updates from Overleaf

\clearpage
\section*{Introduction}

    %\subsection*{1 Why are bottom-up effects important for understanding stability?}
     The plant community is the foundation upon which the myriad species in terrestrial food webs depend for their survival. Disturbances in the plant community have been shown to affect important ecosystem properties such as primary \citep{Hector1999} and secondary production \citep{borer2012plant}, soil respiration and carbon cycling \citep{chen2019plant} and consumer diversity \citep{scherber2010bottom, Baiser2016}.
     The structure of the plant community affects stability across all levels of ecosystem organization, which is crucial for sustaining a healthy energy flow in the ecosystem as a whole~\citep{proulx2010diversity,scherber2010bottom,Rosenblatt2016}. 
     For example, the loss of diversity at the basal level typically causes declines in both abundance and richness of all types of consumers: herbivores, predators, parasitoids, etc. \citep{scherber2010bottom}.

@@

    %\subsection*{5 Motif participation may similarly affect species' particular extinction risk}
    One way to incorporate more nuance into descriptions of how species fit into their communities is to define species' roles based on their participation in different motifs -- unique combinations of small numbers of interacting species~\citep{Stouffer2007,Stouffer2012}. Importantly, these `motif participation' roles capture information about species' direct and indirect interactions, incorporating larger-scale network structure into species-level descriptions~\citep{Cirtwill2015a}. 
    This means that two species with the same in-degrees and trophic levels can still participate in different sets of motifs if, for example, one interacts with specialists and the other interacts with generalists~\citep{Cirtwill2018FoodWebs}. 
    This extra detail may be useful when predicting  extinction risk since, at the network level, the frequencies of different motifs are associated with community stability \citep{prill2005dynamic, bascompte2005simple} and with changes to the composition of plant communities \citep{giling2019plant}. 
    We therefore expect that both the motif structure of a network and species' participation in that structure should be related to risks due to disturbance to basal resources.

@@

@@

	We tested 1) how species persistence varied with motif participation, 2) how motif participation varied with other network properties (in-degree and STL), 3) how network mean persistence varied with network motif profiles, 4) how network motif profiles varied with global structural properties (S and C), and 5) how species persistence and network mean persistence varied with other network properties.
	See Appendix \emph{3.1 - 3.5} for details. 
	All analyses are briefly described below and were performed using the R~\citep{R} packages \emph{lmerTest}~\citep{lmerTest}, \emph{MuMIn}~\citep{MuMIn}, and \emph{vegan}~\citep{vegan}.

@@

        To test if participating more often in a certain motif affected a species' probability of persistence, we fit four linear mixed-effect models (LMMs; one per motif included in the Bayesian networks):
            \begin{equation}
                \Psi_{ink} \approx \rho_{i} + \pi_{k} + \rho_{i}\pi_{k} +
                S_{n}C_{n} + N_n,
                \label{propreq}
            \end{equation}
        \noindent where $\Psi_{ink}$ is the persistence of species $i$, belonging to network $n$, during disturbance level $k$, $\rho_{i}$ is the proportion of the role of species $i$ that is made up by the focal motif, $\pi_k$ is the probability of extinction for a basal resource in disturbance level $k$,  $S_{n}C_{n}$ is a random intercept for the species richness and connectance of network $n$, and $N_n$ is a random effect of network $n$, intended to account for the non-independence of motif participations for species in the same network.
        All predictors were centered and scaled before fitting the models.         
        To test whether these general trends were consistent across networks, we examined the distribution of slopes of simplified regressions fit to a single network and level of disturbance (Appendix \emph{3.1}).

    \subsubsection*{2) Motif participation and other network properties}    
        Different scales of network structure (e.g., local and meso-scale) are not independent. 
        To identify correlations between global network structure and participation in particular motifs, we fit a series of LMs relating motif participation to network size, connectance, and their interaction (Appendix \emph{S3.2}).
        We fit a similar set of  LMMs relating motif participation to in-degree or trophic level (Appendix \emph{S3.2}).
        Taken together, these supplemental models demonstrate how motif participation relates to other measures of network structure.

@@

    In summary, our results show that participating in different three-species motifs is related to a species' risk of extinction following a bottom-up disturbance, and that these relationships synthesize information provided by other, simpler measures of species roles.
    Our results also underscore the importance of considering the level of disturbance when identifying which species are most at risk of extinction.
    There were strong interactions between the level of disturbance and the relationships between persistence and multiple measures of species roles (motifs and in-degree).
    These simulated results echo other warnings about `tipping points' where previously beneficial traits can become risky~\citep{Latty2019,Golubski2016,Tylianakis2014}.
    These disturbance effects mean that efforts to define a set of target species for conservation interactions must take the current (and future) level of disturbance the community experiences into account.

>>>>

\clearpage
\section*{Introduction} %New

    %\subsection*{1 Why are bottom-up effects important for understanding stability?}
     The plant community is the foundation upon which the myriad species in terrestrial food webs depend for their survival. Disturbances in the plant community have been shown to affect important ecosystem properties such as primary \citep{Hector1999} and secondary production \citep{borer2012plant}, soil respiration and carbon cycling \citep{chen2019plant} and consumer diversity \citep{scherber2010bottom, Baiser2016}.
     The structure of the plant community affects stability across all levels of ecosystem organization, which is crucial for sustaining a healthy energy flow in the ecosystem as a whole~\citep{proulx2010diversity, scherber2010bottom, Rosenblatt2016}. 
     For example, the loss of diversity at the basal level typically causes declines in both abundance and richness of all types of consumers: herbivores, predators, parasitoids, etc. \citep{scherber2010bottom}.

    %\subsection*{2 Global and local network properties are known to affect stability}
    How population decline or extinction of one species leads to loss of other species via direct and indirect effects (i.e., secondary extinctions) is a vibrant area of research \citep{Santos2021,curtsdotter2011robustness, dunne2009cascading, Eklof2006}.
    Several global (whole-network) as well as local (single-species) properties are known to influence how species respond to disturbances. 
%network functioning \citep{Petchey2002}, stability \citep{Allesina2012} and robustness to disturbances \citep{Dunne2002, Eklof2006}. 
    Two well-known, influential global properties are network size, the number of species present in the network, and network connectance, the number of realized feeding interactions.
    Based on simulations of random networks, large networks were initially thought to be less stable~\citep{May1972}.
    The presence of large, stable, empirical networks has spurred research into how other structural characteristics, including connectance, may allow large numbers of species to persist~\citep{Dunne2002}.
    Highly-connected networks may be more resistant to secondary extinctions \citep{Dunne2002, Eklof2006} because consumers, on average, have access to a larger number of prey species. However, in a well-connected network a disturbance may spread more easily and cause more secondary extinctions due to the presence of many possible pathways.
    The net effect of connectance on extinction therefore likely depends on the severity of disturbance and the ability of consumers to subsist on a reduced set of prey.

    Local properties known to affect species' response to disturbances include trophic level (a measure of the path length between a consumer and a basal resource) and in-degree (number of prey items of a predator). 
    Species with high trophic levels (long paths to basal resources) and low in-degree (few prey) are generally at higher risks of secondary extinction \citep{binzer2011susceptibility, Eklof2006}. 
    However, in-degree and trophic level are both relatively coarse descriptions of a species' role in its food web~\citep{Cirtwill2018FoodWebs}. For example, a consumer with low trophic level may be at high risk of extinction if it consumes prey which are themselves very likely to be affected by disturbance,and a specialist may not necessarily have a high risk of extinction if it consumes a basal resource which is unlikely to be affected by disturbance.

    %\subsection*{3 Motif participation may similarly affect species' particular extinction risk}
    One way to incorporate more nuance into descriptions of how species fit into their communities, and provide information linking global-scale and local-scale properties, is to define species' roles based on their participation in different motifs -- unique combinations of small numbers of interacting species~\citep{Stouffer2007,Stouffer2012}. 
    Such motif profiles of species are easily calculated once we know the network structure.
    Importantly, these `motif participation' roles capture information about species' direct and indirect interactions, incorporating larger-scale network structure into species-level descriptions~\citep{Cirtwill2015a}. 
    This means that two species with the same in-degrees and trophic levels can still participate in different sets of motifs if, for example, one interacts with specialists and the other interacts with generalists~\citep{Cirtwill2018FoodWebs}.

    This extra detail may be useful when predicting extinction risks of single species, but to our knowledge this has not yet been investigated. 
    However, at the network level, it has been shown that the frequencies of different motifs are associated with community stability \citep{prill2005dynamic, bascompte2005simple} and with changes to the composition of plant communities \citep{giling2019plant}.
    While these studies uncovered the potential influence of network-level motif frequencies on network stability, a disturbance may likely alter the network and the interactions between species in different ways. 
    This can affect the influence of motif frequencies on network stability, as it has been demonstrated that an unstable motif may appear stabilizing dependent on interactions with other species \citep{Kratina2012}.

%\subsection*{Here we test...}
    Therefore, we are interested in whether a consumer's extinction risk depends on 1) the motifs in which the species participates and 2) the motif profile of the full food web, in scenarios with various levels of disturbance targeting the basal level.
    In addition, we want to understand how the information provided by motifs is related to other global and local food web properties in order to better synthesize information provided by different scales of network structure.
    We model extinction probabilities using Bayesian network modeling \citep{Eklof2013, Haussler2020}
    Of the 13 three-species motifs that can occur in empirical food webs, four motifs are over-represented in empirical networks and are thought to contribute to stability; the three-species chain, omnivory, apparent competition, and direct competition \citep{Stouffer2007, Borrelli2015a, giling2019plant}.
    Therefore, we focus on these four particularly important motifs.

     %\subsection*{2 Bayesian networks are a fantastic way to model bottom-up effects}

% OLD INTRODUCTION BELOW
\section*{Introduction} %Old, for comparison

    %\subsection*{1 Why are bottom-up effects important for understanding stability?}
     The plant community is the foundation upon which the myriad species in terrestrial food webs depend for their survival. Disturbances in the plant community have been shown to affect important ecosystem properties such as primary \citep{Hector1999} and secondary production \citep{borer2012plant}, soil respiration and carbon cycling \citep{chen2019plant} and consumer diversity \citep{scherber2010bottom, Baiser2016}.
     The structure of the plant community affects stability across all levels of ecosystem organization, which is crucial for sustaining a healthy energy flow in the ecosystem as a whole~\citep{proulx2010diversity, scherber2010bottom, Rosenblatt2016}. 
     For example, the loss of diversity at the basal level typically causes declines in both abundance and richness of all types of consumers: herbivores, predators, parasitoids, etc. \citep{scherber2010bottom}.

@@

    %\subsection*{5 Motif participation may similarly affect species' particular extinction risk}
    One way to incorporate more nuance into descriptions of how species fit into their communities is to define species' roles based on their participation in different motifs -- unique combinations of small numbers of interacting species~\citep{Stouffer2007,Stouffer2012}. This can be referred to as meso-scale properties and describe how a species is interacting with its closest neighbours. As such, meso-scale properties is thus one a scale between local properties and global properties. Importantly, these `motif participation' roles capture information about species' direct and indirect interactions, incorporating larger-scale network structure into species-level descriptions~\citep{Cirtwill2015a}. 
    This means that two species with the same in-degrees and trophic levels can still participate in different sets of motifs if, for example, one interacts with specialists and the other interacts with generalists~\citep{Cirtwill2018FoodWebs}. 
    This extra detail may be useful when predicting  extinction risk since, at the network level, the frequencies of different motifs are associated with community stability \citep{prill2005dynamic, bascompte2005simple} and with changes to the composition of plant communities \citep{giling2019plant}. 
    We therefore expect that both the motif structure of a network and species' participation in that structure should be related to risks due to disturbance to basal resources.

@@

    \subsection*{Bayesian network framework}	Traditionally, there are two main approaches for studying secondary extinctions. First, there are topological models, based only on food web structure \citep{dunne2009cascading}. Here, extinctions only affect other species in the network once a consumer has lost all of its prey and therefore must go extinct. 
    The second approach uses dynamical models which explicitly simulate population dynamics using a system of differential equations \citep{binzer2011susceptibility}. Dynamical models take changes in prey or predator densities into account when calculating the densities of other species in the network. 
    This additional detail means that realistic processes such as indirect interactions can be included, but also means that dynamical models are more parameter intensive, and simulations are more time-consuming, than topological models.

@@

	We tested 1) how species persistence varied with motif participation, 2) how motif participation varied with other network properties (in-degree and STL), 3) how network mean persistence (the average persistence of consumers) varied with network motif profiles, 4) how network motif profiles varied with global structural properties (S and C), and 5) how species persistence and network mean persistence varied with other network properties (S, C, level of disturbance, and their interactions).
	See Appendix \emph{3.1 - 3.5} for details. 
	All analyses are briefly described below and were performed using the R~\citep{R} packages \emph{lmerTest}~\citep{lmerTest}, \emph{MuMIn}~\citep{MuMIn}, and \emph{vegan}~\citep{vegan}.

@@

        To test if participating more often in a certain motif affected a species' probability of persistence, we fit four linear mixed-effect models (LMMs; one per motif included in the Bayesian networks):
            \begin{equation}
                \Psi_{ink} \approx \rho_{i} + \pi_{k} + \rho_{i}\pi_{k} +
                S_{n}C_{n} + N_n,
                \label{propreq}
            \end{equation}
        \noindent where $\Psi_{ink}$ is the persistence of species $i$, belonging to network $n$, during disturbance level $k$, $\rho_{i}$ is the proportion of the role of species $i$ that is made up by the focal motif, $\pi_k$ is the probability of extinction for a basal resource in disturbance level $k$,  $S_{n}C_{n}$ is a random intercept for the species richness and connectance of network $n$, and $N_n$ is a random effect of network $n$, intended to account for the non-independence of motif participations for species in the same network.
        All predictors were centered and scaled before fitting the models.         
        To test whether these general trends were consistent across networks, we visually examined the distribution of slopes of simplified regressions fit to a single network and level of disturbance (Appendix \emph{3.1}).

    \subsubsection*{2) Motif participation and other network properties}

        Properties which refer to different scales of network structure, including different amounts of direct and indirect interactions with the focal species, are not independent. 
        To identify correlations between global network structure (i.e., the whole network) and participation in particular motifs (a meso-scale description of species roles), we fit a series of LMs relating motif participation to network size, connectance, and their interaction (Appendix \emph{S3.2}).
        We fit a similar set of  LMMs relating motif participation to in-degree or trophic level (i.e., local measures of a species' role; Appendix \emph{S3.2}).
        Taken together, these supplemental models demonstrate how motif participation relates to other measures of network structure.

@@

    In summary, our results show that participating in different three-species motifs is related to a species' risk of extinction following a bottom-up disturbance, and that these relationships synthesize information provided by other, simpler measures of species roles.
    Our results also underscore the importance of considering the level of disturbance when identifying which species are most at risk of extinction.
    There were strong interactions between the level of disturbance and the relationships between persistence and multiple measures of species roles (motifs and in-degree).
    These simulated results echo other warnings about `tipping points' where previously beneficial traits can become risky~\citep{Latty2019,Golubski2016,Tylianakis2014}.
    For example, at low levels of disturbance a high degree may be protective since species can compensate if one or a few prey are lost. 
    At high disturbance, however, a high-degree species is likely to be affected by disturbances through many paths and have lower persistence.
    These disturbance effects mean that efforts to define a set of target species for conservation interactions must take the current (and future) level of disturbance the community experiences into account.
